### 2015-03-14 staticanalysis.tex
split the monolithic technique section into finer grained sections

We have implemented our approach in \whoop, a prototype infrastructure that combines static lockset analysis with state-of-the-art compilation and sequential verification techniques to (i) soundly analyse Linux kernel modules for data races and (ii) accelerate a plugged-in bug finder.

\begin{figure*}
\centering
\includegraphics[width=.99\linewidth]{img/whoop.pdf}
\caption{The \whoop infrastructure, empowered by state-of-the-art compilation (Clang/LLVM and SMACK) and verification (Boogie and \corral) tools}
\label{fig:whoop}
\end{figure*}

Figure~\ref{fig:whoop} depicts the \whoop infrastructure. Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model. The environmental model (i.e. stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed. Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{tech:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs its static lockset analysis (see Section~\ref{tech:analysis}) to detect all potential data races in the abstract program. Finally, \whoop uses information from the static analysis phase to accelerate bug finding with \corral (see Section~\ref{tech:bugfinding}).

\subsection{Translation to Boogie IVL}
\label{tech:translation}

The main aim of this phase is to take an input Linux kernel module written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g. Boogie and \corral). To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.

First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given module, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points. This function acts as the main entry point to the module after the latter has been registered with the kernel. This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{tech:analysis} and~\ref{tech:bugfinding}). The output is a file that contains information regarding the entry points and a file with the rewritten program.

Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. SMACK then translates the LLVM-IR to Boogie IVL. SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL. To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.

\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
static void entrypoint(struct net_device *dev) {
  struct shared *tp = netdev_priv(dev);
  mutex_lock(&tp->wk.mutex);
  tp->field = 4;
  mutex_unlock(&tp->wk.mutex);
}
\end{lstlisting}

Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK. The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource. Notice that functional calls (e.g. to lock and unlock a mutex) are preserved in the translation.

\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
procedure entrypoint(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call mutex_lock($p3);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    $M.0[$p4] := 4;
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call mutex_unlock($p6);
    return;
}
\end{boogie}

\subsection{Static Pairwise Lockset Analysis}
\label{tech:analysis}

We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.

The proof sketch behind our static lockset analysis is given in Section~\ref{proof}. The verification technique is described in more detail in Section~\ref{method}. The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}. Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}. Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.

@@

\subsection{Verification Method}
\label{method}

@@

\subsection{Bug Finding Phase}
\label{tech:bugfinding}

%\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution). Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation. We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious. Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

>>>>

After the original program has been translated to the abstract Boogie program, the core \whoop tool is invoked, which performs \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.

The proof sketch behind our static lockset analysis is given in Section~\ref{proof}. The verification technique is described in more detail in Section~\ref{method}. The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}. Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}.

@@

\subsection{Instrumentation and Analysis}
\label{method}

\whoop Engine is invoked. In this phase, \whoop instruments the program for \emph{static pairwise lockset analysis}

@@

### 2015-03-14 whoop.tex
split the monolithic technique section into finer grained sections

Figure~\ref{fig:whoop} depicts the \whoop infrastructure. Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model. The environmental model (i.e. stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed. Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{tech:translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs its static lockset analysis (see Section~\ref{tech:analysis}) to detect all potential data races in the abstract program. Finally, \whoop uses information from the static analysis phase to accelerate bug finding with \corral (see Section~\ref{tech:bugfinding}).

\subsection{Translation to Boogie IVL}
\label{tech:translation}

The main aim of this phase is to take an input Linux kernel module written in C, together with an environmental model, and output an over-approximated version of the program in Boogie IVL, a simple imperative language with well-defined verification-focused semantics used as the input to a number of cutting-edge verifiers (e.g. Boogie and \corral). To achieve this translation, \whoop uses three LLVM\footnote{http://llvm.org}-based tools: Chauffeur, Clang\footnote{http://clang.llvm.org} and SMACK\footnote{https://github.com/smackers/smack}~\cite{rakamaric2014smack}.

First, \whoop runs Chauffeur, a Clang-frontend that we developed to perform two tasks: (i) visit the abstract syntax tree (AST) of the original program and extract all available entry point identifier names in the given module, together with the identifier names of their corresponding kernel callbacks; and (ii) instrument the original program with a function that calls all available entry points. This function acts as the main entry point to the module after the latter has been registered with the kernel. This simple rewriting does not alter the semantics of the original program, is required so that Clang does not compile away any uncalled entry points (because we closed the environment using stub functions) and is also later used by \whoop to help instrument asynchronous calls to entry points and create checker procedures (see Sections~\ref{tech:analysis} and~\ref{tech:bugfinding}). The output is a file that contains information regarding the entry points and a file with the rewritten program.

Next, the rewritten program is passed to Clang which compiles it to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. SMACK then translates the LLVM-IR to Boogie IVL. SMACK leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of the original program in Boogie IVL. To achieve scalability, SMACK is using a split-memory model instead of a monolithic one that would be difficult for backend verifiers to reason about~\cite{rakamaric2009scalable}.

\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
static void entrypoint(struct net_device *dev) {
  struct shared *tp = netdev_priv(dev);
  mutex_lock(&tp->wk.mutex);
  tp->field = 4;
  mutex_unlock(&tp->wk.mutex);
}
\end{lstlisting}

Example~\ref{fig:original_program} shows a simple program in C and Example~\ref{fig:smack_translation} shows its corresponding translation in Boogie IVL using SMACK. The function \texttt{\$pa} models pointer arithmetic, while the map \texttt{\$M.0} is a memory region that typically denotes a shared resource. Notice that functional calls (e.g. to lock and unlock a mutex) are preserved in the translation.

\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
procedure entrypoint(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call mutex_lock($p3);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    $M.0[$p4] := 4;
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call mutex_unlock($p6);
    return;
}
\end{boogie}

\subsection{Static Pairwise Lockset Analysis}
\label{tech:analysis}

%Next, the Boogie program is transformed into a sequentialised abstract program, using pair-wise sequentialisation and lock set instrumentation as discussed in Section~\ref{method}. Finally the abstract program is send to the Boogie verification engine, which generates verification conditions~\cite{barnett2005weakest} and discharges them to a theorem prover. Successful verification implies that the original driver is free of data races, while an error denotes a \emph{potential} data race.

We now present \emph{static pairwise lockset analysis}, a novel verification technique for proving data race freedom in Linux kernel modules. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing sequential verification techniques.

The proof sketch behind our static lockset analysis is given in Section~\ref{proof}. The verification technique is described in more detail in Section~\ref{method}. The summarisation that we perform to achieve scalability is discussed in Section~\ref{summarisation}. Various optimisations that we perform to achieve precision are described in Section~\ref{optimisations}. Finally, the prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.

\subsection{Static Lockset Analysis Proof Sketch}
\label{proof}

We now formalise our approach to verifying data race freedom in a kernel module by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a module, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this module. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all modules that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking. The actual Linux kernel employs sophisticated fine-grained locking, but this is part of the environment which is abstracted away in our models.

For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$. Furthermore, for each entry point $\mathit{ep}_{i}$, we compute its read-set $R_{i}\lbrack m\rbrack \rightarrow \{true, false\}$ and its write-set $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which map each $m$ to true if and only if $\mathit{ep}_{i}$ reads or writes respectively to $m$ during some execution. In the actual implementation we avoid the use of quantifiers by using a different lockset, read-set and write-set for each shared memory location. \PDComment{any concrete reference that quantifiers might be harder for SMT solving?}

\begin{theorem}
\label{theorem:locksets}
For each pair of entry points $\mathit{ep}_{i}, \mathit{ep}_{j}\in \{\mathit{ep}_{1}, \mathit{ep}_{2}, ..., \mathit{ep}_{n}\}$, where $i$ may be equal to $j$, and for each memory location $m$, if $(W_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \wedge (W_{i}\lbrack m\rbrack \vee R_{j}\lbrack m\rbrack) \wedge (R_{i}\lbrack m\rbrack \vee W_{j}\lbrack m\rbrack) \implies (\mathit{LS}_{i}\lbrack m\rbrack \cap \mathit{LS}_{j}\lbrack m\rbrack \not= \varnothing)$, then the kernel module with entry points $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ is free from data races.
\end{theorem}

A sketch of how the above theorem can be proved is as follows. Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$. By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$. By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both $ep_1$ and $ep_2$. As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

\subsection{Verification Method}
\label{method}

\begin{table*}
\small
\begin{center}
\begin{tabular}{lll}
\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
\noalign{\smallskip}\toprule
\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
\noalign{\smallskip}
                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
\noalign{\smallskip}
                                       & W$\lbrack $ptr$\rbrack := true;$ &
\end{tabular}
\end{center}
\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
\label{tab:instrumentation}
\end{table*}

Our verification technique begins by performing \emph{two-thread reduction}, a sound abstraction that removes all but two arbitrary threads, each running an entry point of the originally concurrent program, and then performs \emph{pairwise sequentialisation}, which combines the two arbitrary threads in a single sequential pair. This is achieved by replacing the \texttt{return} statement of the first entry point with a \texttt{goto} statement that jumps execution to the first statement of the second entry point. This process repeats until all pairs of entry points have been sequentialised. To achieve soundness, each time an entry point performs a read access to a shared resource, we return a nondeterministic value. This over-approximates any effects from all the unmodeled threads on the device driver shared state.

Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point. The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}. The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger. This whole process is summarised below:

\begin{description}
  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false. Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty. Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$. This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.

  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger. Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty. Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty. If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$. In this case a counterexample is generated and reported to the user. To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
\end{description}

\subsection{Watchdog Summarisation}
\label{summarisation}

Draft.

\subsection{Optimisations}
\label{optimisations}

Draft.

\subsection{Assumptions and Limitations}
\label{limitations}

\subsection{Bug Finding Phase}
\label{tech:bugfinding}

%\textsc{Whoop} is currently working on an unmodified network device driver (the RealTek r8169 which is part of the official Linux kernel distribution). Our immediate future work includes applying invariant generation to tame the effect of our coarse over-approximation. We also plan to introduce counterexample feasibility checking to evaluate if a reported bug is real or spurious. Towards this, we plan to feed a counterexample generated by \textsc{Whoop} into a precise bug finder, such as Corral~\cite{lal2012corral}, which will be guided only to execution paths that may manifest the bug, to either confirm it or discard it as a false positive.

>>>>

Figure~\ref{fig:whoop} depicts the \whoop infrastructure. Initially, \whoop accepts a kernel module written in C, together with a Linux environmental model. The environmental model (i.e. stub Linux header files) is required to ``close'' the module and allow it to be translated and subsequently analysed. Both the module and the environmental model are passed to the translation phase of \whoop (see Section~\ref{translation}), which outputs an over-approximated program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}. Next, \whoop performs its static lockset analysis (see Section~\ref{staticanalysis}) to detect all potential data races in the abstract program. Finally, \whoop uses information from the static analysis phase to accelerate bug finding with \corral (see Section~\ref{bugfinding}).
